\documentclass[../main.tex]{subfiles}
\graphicspath{{../images/}}

\begin{document}
\subsection*{Lecture 9: \hfill  2/5/24}
\hrule \vspace{10px}
\section{Calculus of Variations}

\paragraph*{Why do we care?}
\begin{itemize}
    \item What is the shortest distance between two points in a 2D plane?
    \item What is the shortest path between two points on a sphere?
    \item What is the fastest path for a ball to roll down a hill?
    \item For a car driving on a flat path $A \to B$, what shape of a pot hole will minimize
    the time it takes to get from $A \to B$?
\end{itemize}

For some path $a \to b$, we have a path defined as an integral
\begin{align*}
    S = \int_a^b f(x, y, y') \dd x
\end{align*}
with a \emph{Goal}: find $y(x)$ that minimizes $S$ (path).

\paragraph*{Path Length:}
\begin{align*}
    l = \int_a^b \sqrt{\dd x^2 + \dd y^2} = \int_a^b \sqrt{1 + y'^2} \dd x
\end{align*}
where $y' = \frac{\dd y}{\dd x}$. To minimize $y = f(x)$ it is equivalent to finding where
\begin{align*}
    f'(x) = 0
\end{align*}
where we note that this could be a maximium point, but it is usually a minimum in these cases.
Another look at this function:
\begin{align*}
    f'(x) = \frac{\dd y}{\dd x} = \lim_{\Delta x \to 0} \frac{y(x + \Delta x) - y(x)}{\Delta x}
\end{align*}
we can define a small change in the path $y(x)$ as
\begin{align*}
    y(x) + \delta y(x)
\end{align*}
where 
\begin{align*}
    \delta y(x_2) = 0 \quad \delta y(x_2) = 0
\end{align*}
so the change in the path is
\begin{align*}
    \delta S = \int_a^b \delta f \dd x
\end{align*}
and from the change of variables
\begin{align*}
    \delta f = \pdv{f}{y} \delta y + \pdv{f}{y'} \delta y' \qquad \delta y' = \dv{x} \delta y
\end{align*}
thus we have
\begin{align*}
    \delta S = \int_a^b \qt( \pdv{f}{y} \delta y + \pdv{f}{y'} \dv{x} \delta y) \dd x
\end{align*}
this is the line integral of the change in the new path
\begin{align*}
    \delta S = S_{new} - S_{old}
\end{align*}
looking at the second term: using integration by parts
\begin{align*}
    \int_a^b \qt(\pdv{f}{y} \dv{x} \delta y) \dd x = \qt[\pdv{f}{y'} \delta y]\eval_a^b
        - \int_a^b \qt(\dv{x} \pdv{f}{y'}) \delta y \dd x 
\end{align*}
the first term is zero because $\delta y(a) = \delta y(b) = 0$. Thus we have
\begin{align*}
    \delta S = \int_a^b \qt[\pdv{f}{y} - \qt(\dv{x} \pdv{f}{y'})] \delta y \dd x
\end{align*}
Near a minimum, $\delta S = 0$ for any small $\delta y$. So the the terms in the brackets must be
zero as well! This gives us the \textbf{Euler-Lagrange Equation}:
\begin{align*}
    \pdv{f}{y} - \dv{x} \pdv{f}{y'} = 0
\end{align*}
\paragraph*{NOTE:} $\delta S$ is the variation of $S$(some number) under $\delta y$(a function).

\paragraph*{Example:} Shortest path between two points $a \to b$ in a 2D cartesian plane.

\subparagraph*{} Goal: find $y(x)$ tha minimzes the path length $l = \int_a^b \sqrt{1 + y'^2} \dd x$
where $f(x, y, y') = \sqrt{1 + y'^2}$.
\begin{align*}
    \pdv{f}{y} &= 0 \\
    \pdv{f}{y'} &= \frac{2y'}{2\sqrt{1 + y'^2}} = \frac{y'}{\sqrt{1 + y'^2}} \\
\end{align*}
From the EL:
\begin{align*}
    \dv{x}(\frac{y'}{\sqrt{1 + y'^2}}) = \pdv{f}{y} =  0 \\
\end{align*}
and
\begin{align*}
    \frac{y'}{\sqrt{1 + y'^2}} &= Const = C \\
    y'^2 &= C (1 + y'^2) \\
    y'^2 &= \frac{C}{1 - C} \\
    y' &= \pm \sqrt{\frac{C}{1 - C}} = \pm k \\
    y &= \pm kx + b
\end{align*}
which is just a straight line as we expected. 

\paragraph*{Example:} The Brachistochrone. 
\subparagraph*{} Goal: Find $y(x)$ that minimizes $t = \int_a^b \dd t$ where 
\begin{align*}
    t = \frac{s}{v} \to \dd t = \frac{\dd s}{v}
\end{align*}
and $v$ is the velocity which can be found using the conservation of energy:
\begin{align*}
    \frac{1}{2} m v^2 = mg(y_o - y) \to v = \sqrt{2g(y_o - y)}
\end{align*}
thus we have
\begin{align*}
    \dd t = \frac{\dd s}{v} = \frac{\sqrt{1 + y'^2}}{\sqrt{2g(y_o - y)}} \dd x
\end{align*}
where $f(x, y, y') = \frac{\sqrt{1 + y'^2}}{\sqrt{2g(y_o - y)}}$. Using EL:
\begin{align*}
    \pdv{f}{y'} &= \frac{y'}{\sqrt{1 + y'^2}\sqrt{2g(y_o - y)}} \\
    \pdv{f}{y} &= \frac{\sqrt{1 + y^2} \sqrt{2g}}{(2g(y_o - y))^{3/2}} 
\end{align*}
From the initial conditions: $x = y = 0, \quad \dot x = \dot y = 0$ (assuming no friction). So
\begin{align*}
    f = \frac{\sqrt{1 + y'^2}}{\sqrt{2gy}} \to \pdv{f}{y} = \frac{\sqrt{1 + y'^2}}{\sqrt{2gy^3}}
\end{align*} 
and if $y(x)$ satisfies EL, then we have a conserved quantity (prove in HW):
\begin{align*}
    f - y' \pdv{f}{y'} = const
\end{align*}
we will soon find out that this is the energy of the system\dots
we can solve for $y(x)$ using the conserved quantity:
\begin{align*}
    \pdv{f}{y'} = \frac{1}{\sqrt{2gy}} \frac{y'}{\sqrt{1 + y'^2}}
\end{align*}
so 
\begin{align*}
    f - y' \pdv{f}{y'} &= C \\
    \frac{1}{\sqrt{2gy}}\qt(\sqrt{1 + y'^2} - \frac{y'^2}{\sqrt{1 + y'^2}}) &= C  \\
    \frac{1}{\sqrt{2gy}\sqrt{1 + y'^2}} (1 + y'^2 - y'^2) &= \sqrt{\frac{1}{4ga}}
\end{align*}
using
\begin{align*}
    \frac{1}{y(1 + y'^2)} &= \frac{1}{2a} \implies 1 + y'^2 = \frac{2a}{y} \implies y'^2 = \frac{2a}{y} - 1
\end{align*}
this can be solved using separation of variables:
\begin{align*}
    \dv{y}{x} = \sqrt{\frac{2a}{y} - 1} \implies \dd x = \dd y \sqrt{\frac{y}{2a - y}}
\end{align*}
and integration both sides:
\begin{align*}
    \int_a^b \dd y \sqrt{\frac{y}{2a - y}} = x - x_a = x
\end{align*}
using a change of variables $y = a (1 - \cos \theta); \quad \dd y = a \sin \theta \dd \theta$ and 
a substitution $\sin\theta = \sqrt{1 - \cos^2 \theta} = \sqrt{(1 - \cos\theta)(1 + \cos\theta)}$:
\begin{align*}
    \int_a^b a \sin \theta \dd \theta \sqrt{\frac{a(1 - \cos \theta)}{a (1 + \cos \theta)}}
    = \int_a^b a(1 - \cos \theta) \dd \theta = a \theta - a \sin \theta
\end{align*}
this is a parametric equation:
\begin{align*}
    x &= a(\theta - \sin \theta) = x(\theta)\\
    y &= a(1 - \cos \theta) = y(\theta)
\end{align*}
where $\theta = \omega t$. This is a curve traced by a point on a wheel AKA cycloid. When we choose
a variable time we get
\begin{align*}
    x(t) &= a(\omega t - \sin \omega t) \\
    y(t) &= a(1 - \cos \omega t)
\end{align*}
and thus we get $\omega = \sqrt{\frac{g}{a}}$. To find $a$ we use the coordiate of the lower second point to find
the curve that goes through the two points.

\paragraph*{Example:} Find two functions $x(u)$, $y(u)$ where the path
\begin{align*}
    S = \int_a^b f(x, x', y, y', u) \dd u 
\end{align*}
is minimized/stationary. We will get two EL equations:
\begin{align*}
    \pdv{f}{x} - \dv{u} \pdv{f}{x'} &= 0 \\
    \pdv{f}{y} - \dv{u} \pdv{f}{y'} &= 0
\end{align*}
e.g. for a distance between two points:
\begin{align*}
    L = \int_a^b \sqrt{\dd x^2 + \dd y^2} = \int_a^b \sqrt{x'^2 + y'^2} \dd u
    \qusing \dd y = \dv{y}{u} \dd u = y' \dd u
\end{align*}
and from the EL equations:
\begin{align*}
    \dv{u} \pdv{f}{x'} = 0 = \dv{u}(\frac{x'}{\sqrt{x'^2 + y'^2}}) \\
    \implies C_1 = \frac{x'}{\sqrt{x'^2 + y'^2}} \quad C_2 = \frac{y'}{\sqrt{x'^2 + y'^2}}
\end{align*}
this also tells us that 
\begin{align*}
    \frac{y'}{x'} = const = \dv{y}{x}
\end{align*}
For $N$ unknown functions in time $t$:
\begin{align*}
    S = \int_a^b f(x_1, x_1', \dots, x_N, x_N', u) \dd u
\end{align*}
where $f$ has $2N + 1$ variables.

\paragraph*{Generalized Coordinates:} $q_1, q_2, \dots, q_N$ we would define the Lagrangian
\begin{align*}
    \mathcal{L} (q_1, \dots, q_N, \dot q_1, \dots, \dot q_N, t)
\end{align*}
and minimize the action
\begin{align*}
    S = \int \mathcal{L} \dd t
\end{align*}
and $N$ EL equations gives the trajectory for the path of minimal action.
\end{document}